---
title: "Portfolio Optimization"
output: rmarkdown::github_document
---

Load libraries
```{r, warning=FALSE, message=FALSE}
library(rvest) # web scraping
library(quantmod) # donwload stock data
library(dplyr) # data manipulation
library(TTR) # calculate returns
library(aTSA) # test arch effects
library(fDMA)# test arch effects
library(stringr) # string manipulation
library(fGarch) # garchFit
library(tidyr) # data manipulation
library(gurobi) # optimization
library(ggplot2) # plots
library(gridExtra) # plots in same grid
library(scales) # plot axis labels
filter <- dplyr::filter
select <- dplyr::select
```

Scrape stock symbols in DJIA from wikipedia.com
```{r}
url <- 'https://en.wikipedia.org/wiki/Dow_Jones_Industrial_Average'
# read from wiki page
tbl <- read_html(url) %>% 
  html_nodes(css = 'table')
# convert to df
tbl <- tbl[2] %>% html_table() %>% as.data.frame()
```

Download stock data from yahoo.com
```{r, warning=FALSE, message=FALSE}
# create new environment for stock data
quotes <- new.env()
# download stock data to quotes environment
getSymbols(tbl$Symbol, src = 'yahoo', from = '2015-01-01', env = quotes)
```

Create xts object with closing prices of the 30 stocks in DJIA
```{r}
# store stock data in quotes enirnoment into a list
q_list <- eapply(quotes, "[") 

# store the 1st stock's closing price in stocks
stocks <- q_list[[1]][,4] # 4th column is closing price

# 2nd to 30th stock in the list and just keep closing price. 
for (i in 2:length(q_list)) {
  stocks <- cbind(stocks, q_list[[i]][,4])
}

# subset to keep specified data range
stocks2 <- stocks["2017-02-01/2019-02-08"]

# saveRDS(stocks, "../data/stocks.RDS")
# stocks <- readRDS("../data/stocks.RDS")
```

Convert prices to returns
```{r}
stocks_r <- stocks2

# for each stock replace prices with returns (percent changes)
for(stock in names(stocks_r) ){
  stocks_r[,stock] <- periodReturn(stocks2[,stock], 
                                   period = "daily", type='arithmetic')
}

# drop fist row bc its NA from returns calculation
stocks_r <- stocks_r[2:nrow(stocks_r)]

# stocks_r2 <- stocks_r
# names(stocks_r2) <- names(stocks_r2) %>% str_replace_all("\\.", "_")
# write.csv(stocks_r2, "../data/stocks_r.csv")
```

ARCH effect test for the stocks
```{r}
# Test if the volatility is predictable, ie autocorrelations exits
# Null Hypothesis: all coefficients in a ARCH model is equal zero

# Arch test for the first stock 
# arch.test expects an arima object
# function doesn't allow you to choose the lag
# first row from arch.test contains results from testing lag 1-4
arch_test <- arch.test(arima(stocks_r[,1], order = c(0,0,0)), output = F) 
colnames(arch_test) <- c("order", "PQ_Stat", "p_value_PQ", 
                         "LM_Stat", "p_value_LM") 
arch_test <- arch_test %>% 
  as.data.frame() %>% slice(1)

# loop through the rest of the stocks and store results from arch effect test
# into a DF
for(i in 2:ncol(stocks_r)){
arch_test_temp <- arch.test(arima(stocks_r[,i], order = c(0,0,0)), 
                            output = F) 

colnames(arch_test_temp) <- c("order", "PQ_Stat", "p_value_PQ", 
                           "LM_Stat", "p_value_LM")
arch_test_temp <- arch_test_temp %>% 
  as.data.frame() %>% slice(1)

arch_test <- bind_rows(arch_test, arch_test_temp)
}

# add a column for the name of the stock associated with the test result
arch_test$stock <- names(stocks_r)

# Arrange by p value, smallest first
arch_test <- arch_test %>% arrange(desc(LM_Stat))

# results of LM test is inconsistent with PQ test. They're inversely related.
# Also different than SAS. maybe something wrong with the library
arch_test
```

ARCH effect test for the stocks using fDMA package
```{r}
p_values <- c()
stats <- c()
# loop through stocks test for arch effect on first lag. 
# save p value in a vector
for(i in 1:ncol(stocks_r)){
  temp <- archtest(stocks_r[,i] %>% as.numeric(), lag =1)
  p_values <- c(p_values, temp$p.value)
  stats <- c(stats, temp$statistic)
}
# create a data frame with the p-value and associated stock name
arch_test2 <- data.frame(stock = names(stocks_r), p_value = p_values,
                         test_stat = stats) %>%
  mutate(stock = as.character(stock))
# arrange by p-value
arch_test2 <- arch_test2 %>% arrange(p_values)

# results are different from the aTSA package. checked p-values in SAS for top 
# 5 ones with lowest p-value using fDMA and they are all below .0001
# will use these results bc it's testing the first lag rather than 1-4 and is 
# consistent with SAS
arch_test2
```
Select top 5 stocks based on smallest p value from LM test
```{r}
# create vector of names of top 5 stocks
top5_names <- arch_test2 %>% slice(1:5) %>% pull(stock)
# select top 5 stocks  
top5_stocks_r <- stocks_r[,top5_names]
```

Fit Garch Models
```{r echo = T, results = 'hide'}
# GARCH model with 1 AR and 1 MA term is typically best. The underlying dist
# is changed to find a better model

# create a data frame with specified cols for recording stats
fit_stat_df <- data.frame(AIC = NA, BIC = NA, SIC = NA, HQIC = NA, 
                          loglikelihood = NA, model = NA, stock = NA, 
                          alpha = NA, beta = NA)
# distribution types (normal, t, sym normal, sym t)
dists <- c('norm', 'std', 'snorm', 'sstd')

# create 4 types of garch(1,1) modeles for each stock and save results in a df
for(stock in names(top5_stocks_r)){
  for(dist in dists) {
    # fit garch model with specified dist
    fit <- garchFit(formula= ~ garch(1,1), data=top5_stocks_r[,stock],
                    cond.dist=dist, include.mean = FALSE)
    # save fit stats (AIC, SBC etc) from model
    fit_stat <- fit@fit$ics %>% as.matrix() %>% t() %>% as.data.frame() %>%
      mutate(loglikelihood = fit@fit$value*-1 %>% unname, model = dist, 
             stock = stock, 
             alpha = fit@fit$par['alpha1'] %>% unname,
             beta = fit@fit$par['beta1'] %>% unname)
    # add fit stats to df
    fit_stat_df <- bind_rows(fit_stat_df, fit_stat)
  }
}
```

Best model for each stock 
```{r}
# Best model for each stock in terms of lowest AIC
fit_stat_df2 <- fit_stat_df %>% slice(-1) %>%
  group_by(stock) %>%
  arrange(AIC) %>%
  mutate(rn = row_number()) %>%
  dplyr::filter(rn == 1) %>%
  dplyr::ungroup()

fit_stat_df2
```

Plots for parameters / stats of each stock
```{r}
# format model types and stock symbols
fit_stat_df2 <- fit_stat_df2 %>% 
  mutate(Model = ifelse(model == "std", "t-Garch", "QGarch-t"),
         stock = str_replace_all(stock, "\\..+", "")) 

# alpha
a_p <- ggplot(fit_stat_df2, aes(x = reorder(stock, -alpha), y = alpha, fill = Model)) +
  geom_col() +
  scale_fill_manual(values=c("#F7AD50", "#3F97D0")) +
  labs(title = "", x = "", y = "Alpha") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black"),
        legend.position = "none") 

# Beta
b_p <- ggplot(fit_stat_df2, aes(x = reorder(stock, -beta), y = beta, fill = Model)) +
  geom_col() +
  scale_fill_manual(values=c("#F7AD50", "#3F97D0")) +
  labs(title = "", x = "", y = "Beta") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black"),
        legend.position = "none")

# loglilihood
ll_p <- ggplot(fit_stat_df2, aes(x = reorder(stock, -loglikelihood), 
                         y = loglikelihood, fill = Model)) +
  geom_col() +
  scale_fill_manual(values=c("#F7AD50", "#3F97D0")) +
  labs(title = "", x = "", y = "Log Likelihood") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black"),
        legend.position = "none") 

grid.arrange(a_p, b_p, ncol=1)  
```


Fit best model for each stock
```{r echo = T, results = 'hide'}

# fit best model for each stock
XOM_fit <- garchFit(formula= ~ garch(1,1), data=top5_stocks_r[,'XOM.Close'],
                    cond.dist="std", include.mean = FALSE)

UTX_fit <- garchFit(formula= ~ garch(1,1), data=stocks_r[,'UTX.Close'],
                    cond.dist="sstd", include.mean = FALSE)

MRK_fit <- garchFit(formula= ~ garch(1,1), data=stocks_r[,'MRK.Close'],
                    cond.dist="std", include.mean = FALSE)

CVX_fit <- garchFit(formula= ~ garch(1,1), data=stocks_r[,'CVX.Close'],
                    cond.dist="sstd", include.mean = FALSE)

MSFT_fit <- garchFit(formula= ~ garch(1,1), data=stocks_r[,'MSFT.Close'],
                    cond.dist="sstd", include.mean = FALSE)

```

5 day volatility forecast
```{r}
# create df with 5 day forecast of volatitlity for each stock
sd_forecast <- data.frame(XOM = predict(XOM_fit, n.ahead = 5) %>%   
                            pull(standardDeviation),
                          UTX = predict(UTX_fit, n.ahead = 5) %>% 
                            pull(standardDeviation),
                          MRK = predict(MRK_fit, n.ahead = 5) %>% 
                            pull(standardDeviation),
                          CVX = predict(CVX_fit, n.ahead = 5) %>% 
                            pull(standardDeviation),
                          MSFT = predict(MSFT_fit, n.ahead = 5) %>% 
                            pull(standardDeviation))
```
5 day volatility forecast plot
```{r}
sd_forecast2 <- sd_forecast
sd_forecast2$Date <- seq(from = as.Date("2019/02/11"), 
                         to =as.Date("2019/02/15"), "days")

sd_forecast2 <- sd_forecast2 %>% gather(key = Stock, value = SD, -Date)

ggplot(sd_forecast2, aes(x = Date, y = SD, color = Stock)) +
  geom_line(size = 1) +
  scale_color_manual(values=c("#F7AD50", "#3F97D0", "tomato1", 
                             "olivedrab2", "#F26419", "#999999")) +
  labs(title = "Forecasted Volatility",  x = "Date (2019)", 
       y = "Volatility (SD)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black"),
        legend.title = element_text(face = "bold")) +
  scale_y_continuous(labels = comma)
```

Median forecasted variance and returns
```{r}
# compute forecasted variance
var_forecast <- sd_forecast * sd_forecast

# df of median forecasted variance for each stock
med_var <- apply(var_forecast,2, median) %>% 
  as.matrix() %>% t() %>% as.data.frame()

# df of median returns for each stock
med_r <- top5_stocks_r %>% apply(2,median) %>% 
  as.matrix() %>% t() %>% as.data.frame()
# remove .Close from column names
names(med_r) <- names(med_r) %>% str_replace(".Close", "")

# df of median forecasted variance and historical median returns
var_r <- bind_rows(med_var, med_r)
var_r$measure <- c('variance', 'return')

# reorder columns to match column order in top_5_stocks_r
var_r <- var_r %>%
       select(names(top5_stocks_r) %>% 
                str_replace("\\..+", ""), measure)

# var_r %>% gather(key = stock, value = v, -measure)
```


Covariance matrix with forecasted variance
```{r }
# create covarinace matrix of historical returns of top 5 stocks
cov_mtx <- cov(top5_stocks_r)
# replace diangonal entries of covariance mtx with zero
diag(cov_mtx) <- 0
# create diagonal matrix with forecasted median variance
var_mtx <- diag(var_r %>% filter(measure == "variance") %>% select(-measure))
# replace diagonal entries of covariance matrix with forecasted median variance
# This replaces historical variance with forecasted variance / volatility
cov_mtx <- cov_mtx + var_mtx 
```

```{r}
saveRDS(returns, "../data/returns.rds")
saveRDS(cov_mtx, "../data/cov_mtx.rds")
```



Optimize portfolio
```{r}
# mean returns
returns = var_r %>% filter(measure == "return") %>% 
  select(-measure) %>% as.numeric()

model <- list()

# Contraint mtx: 
# sum of weights = 1
# weighted sum of returns >= 0.0005 (min .05% return per day)
# weight >= 0 (can't be negative)
model$A <- matrix(c(rep(1,ncol(cov_mtx)),
                  returns),
                  nrow=2,byrow=T)
model$A <- rbind(model$A, diag(rep(1, ncol(cov_mtx))))
model$sense <- c('=', '>=', rep(">=", ncol(cov_mtx)))
model$rhs   <- c(1, .0005, rep(0, ncol(cov_mtx)))

# Objective function. minimize risk. aka covariance of weighed sum of stocks
# cov(p1r1 + ... + p5r5) where p are weights and r are stock returns
model$Q     <- cov_mtx
model$obj   <- c(rep(0,ncol(cov_mtx))) # coeficients for linear term
model$modelsense <- "min"

# run optimization
result <- gurobi(model, list())
result$status

# proportion to put in each stock
result_names <- c(names(top5_stocks_r))
names(result$x) <- result_names
opt_weights <- result$x %>% as.matrix()

# minimum risk (std deviation)
result$objval %>% sqrt()
# sqrt of covariance of weighted sum (risk)
t(opt_weights) %*% cov_mtx %*% opt_weights %>% sqrt()

# expected return
sum(result$x * returns)
```

Portfolio weights pie chart
```{r}
weights_df <- data.frame(Stock = rownames(opt_weights), 
                         Weights = opt_weights * 100) %>%
  mutate(Stock = str_replace_all(Stock, "\\..+", ""))

 ggplot(weights_df, aes(x = "", y = Weights, fill = Stock)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  geom_text(aes(label = paste0(round(weights_df$Weights,2), "%")), 
            position = position_stack(vjust = .5)) +
  scale_fill_manual(values=c("#F7AD50", "#3F97D0", "tomato1", 
                             "olivedrab2", "#F26419", "#999999")) +
  labs(x = NULL, y = NULL, fill = NULL, title = "Portfolio Allocation") + 
  theme_void() + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = .5, size = 22, face = "bold"))
```

Efficient Frontier Calculation
```{r echo = T, results = 'hide'}
# minimum risk (variance / volatility) for given level of min return

# vary the minimum return
min_returns <- c(seq(0.00065, .0013, .0001))
# constraint operators. 
# return is a binding constraint so use = instead of using >= as before
model$sense <- c('=', '=', rep(">=", ncol(cov_mtx)))


ef_return = vector(length = length(min_returns))
ef_risk = min_returns

# minimize risk for given minimum return
for (i in 1:length(min_returns))
{
  model$rhs <- c(1, min_returns[i], rep(0,ncol(cov_mtx)))
  result2 <- gurobi(model, list())
  ef_return[i] = sum(result2$x * returns)
  ef_risk[i] = sqrt(result2$objval)
  
}
```

Efficient frontier plot
```{r}
ef <- data.frame(risk = ef_risk, return = ef_return*100)

ggplot(ef, aes(x = risk, y = return)) +
  geom_line(color = "#F7AD50", size = 1.5) +
  labs(title = "Efficient Frontier",
       x = "Risk (SD)",
       y = "Return (%)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black")) +
  scale_x_continuous( breaks = c(seq(.007,.014, .0005))) +
  scale_y_continuous( breaks = c(seq(.06,.14, .01))) 
```

actual 5 day return from portfolio
```{r}
# 5 days of prices for the 5 stocks in portfolio
stocks_5day <- stocks["2019-02-11/2019-02-15", names(top5_stocks_r)]
day1 <- stocks_5day %>% head(1) %>% as.numeric()
day5 <- stocks_5day %>% tail(1) %>% as.numeric()
# 5 day return
return_5day <- (day5 - day1)/day1

#  5 day return from optimized portfolio
weighted_return <- sum(opt_weights * return_5day) * 100

#  5 day return from baseline portfolio of having equal weight in each stock
equal_weights <- c(rep(.2,5)) %>% as.matrix()
weighted_return_base <- sum(equal_weights * return_5day) *100
```

Portfolio growth over 5 days
```{r}
# number of each stock to purchase. assuming investment = 100k and u can buy 
# partial stocks (for illustration purposes)
allocation <-(t(opt_weights) * 100000) / (stocks_5day %>% 
                                            head(1)) %>% as.numeric() 
allocation <- matrix(rep(allocation, 5),  nrow = 5, byrow = T)
# value of each stock holding
portfolio_value <- allocation * stocks_5day %>% as.data.frame()
# total portfolio value
portfolio_value$total = rowSums(portfolio_value)


# number of each stock to purchase for baseline portfolio. 
# assuming investment = 100k and u can buy partial stocks (for illustration purposes)
allocation_base <-(t(equal_weights) * 100000) / (stocks_5day %>% 
                                            head(1)) %>% as.numeric()
allocation_base <- matrix(rep(allocation_base, 5),  nrow = 5, byrow = T)
# value of each stock holding
portfolio_value_base <- allocation_base * stocks_5day %>% as.data.frame()
# total portfolio value
portfolio_value_base$total = rowSums(portfolio_value_base)
```

Portfolio growth chart
```{r}
portfolios_5day <- data.frame(Date = rownames(portfolio_value),
                              Optimized = portfolio_value$total,
                              Baseline = portfolio_value_base$total) %>%
  mutate(Date = as.POSIXct(Date))

portfolios_5day <- portfolios_5day %>% 
  gather(key = Portfolio, value = Amount, -Date)

ggplot(portfolios_5day, aes(x = Date, y = Amount, color = Portfolio)) +
  geom_line(size = 1.3) +
  scale_colour_manual(values=c( "#3F97D0", "#F7AD50")) +
  labs(title = "Portfolio Growth",  x = "Date (2019)", 
       y = "Total Amount ($)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black"),
        legend.position = c(0.9, 0.2),
        legend.title = element_text(face = "bold")) +
  scale_y_continuous(labels = comma) 
  
```


actual 5 day risk from portfolio
```{r}
# 6 days of price for the 5 stocks in portfolio
stocks_5day_r <- stocks["2019-02-8/2019-02-15", names(top5_stocks_r)]
# covert prices to returns
for(stock in names(stocks_5day_r)){
  stocks_5day_r[,stock] <- periodReturn(stocks_5day_r[,stock], period = "daily",
                                type='arithmetic')
}
# drop the first row
stocks_5day_r <- stocks_5day_r %>% tail(5)
# covariance matrix of 5 days of returns
cov_mtx_5day <- cov(stocks_5day_r)

# covariance of weighted sum of returns = risk 
# cov(p1r1 + ... + p5r5) where p are weights and r are returns
risk_5day <- t(opt_weights) %*% cov_mtx_5day %*% opt_weights %>% sqrt()
risk_5day_base <- t(equal_weights) %*% cov_mtx_5day %*% equal_weights %>% sqrt()

```

Plots of actual 5 day risk and return
```{r}
return_5day_df <- data.frame(return = c(weighted_return, weighted_return_base),
                             portfolio = c("Optimized", "Baseline"))

risk_5day_df <- data.frame(risk = c(risk_5day , risk_5day_base), 
                           portfolio = c("Optimized", "Baseline"))

p1 <- ggplot(return_5day_df, aes(x = portfolio, y = return)) +
  geom_col(fill = "#F7AD50", color = '#3F97D0') +
  labs(
       y = "Return (%)",
       x = '') +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black")) +
  scale_y_continuous( breaks = c(seq(0,4, .75)))

p2 <- ggplot(risk_5day_df, aes(x = portfolio, y = risk)) +
  geom_col(fill = "#F7AD50", color = '#3F97D0') +
  labs(y = "Risk (SD)", x = "") +
  theme_bw() +
  theme(plot.title = element_text(hjust = .5, size = 22, face = "bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=16),
        axis.line = element_line(colour = "black")) 

grid.arrange(p1, p2, ncol = 2)
```




